---
title: "Support Vector Machines (SVM)"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center"
)
```

```{r libraries}
# ── Install any missing packages before loading ──────────────────────────────
required_pkgs <- c("caret", "e1071", "ggplot2", "dplyr", "tidyr",
                   "gridExtra", "kernlab", "MASS", "RColorBrewer")

new_pkgs <- required_pkgs[!required_pkgs %in% installed.packages()[,"Package"]]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cloud.r-project.org")

# ── Load libraries ────────────────────────────────────────────────────────────
library(caret)       # unified ML interface
library(e1071)       # SVM implementation (svm())
library(ggplot2)     # elegant graphics
library(dplyr)       # data manipulation
library(tidyr)       # data reshaping
library(gridExtra)   # arrange multiple ggplots
library(kernlab)     # kernel-based ML (ksvm)
library(MASS)        # mvrnorm – multivariate normal data generation
library(RColorBrewer)

set.seed(42)         # reproducibility
```

---

# What Is a Support Vector Machine?

A **Support Vector Machine (SVM)** is a supervised machine-learning algorithm used
for **classification** and **regression**.  Its core idea is elegantly geometric:

> *Find the hyperplane that separates two classes while maximising the gap
> (margin) between them.*

## Key Intuition

Imagine plotting two groups of points on a piece of paper.  Many straight lines
could separate them – SVM finds the **one line** that is as far as possible from
both groups.  That distance is the **margin**, and the points sitting on its edge
are called **support vectors** (they literally "support" the boundary).

```{r intuition-plot, fig.cap="Fig 1 – The SVM margin intuition with two linearly separable classes."}
# ── Generate a clean, linearly-separable toy dataset ─────────────────────────
gen_sep <- function(n = 30, seed = 1) {
  set.seed(seed)
  cls_A <- MASS::mvrnorm(n, mu = c(-2, -2), Sigma = diag(0.6, 2))
  cls_B <- MASS::mvrnorm(n, mu = c( 2,  2), Sigma = diag(0.6, 2))
  df <- data.frame(
    x1    = c(cls_A[,1], cls_B[,1]),
    x2    = c(cls_A[,2], cls_B[,2]),
    class = factor(rep(c("Class A", "Class B"), each = n))
  )
  df
}

toy <- gen_sep()

# Fit a linear SVM and extract the decision boundary manually
mdl_lin <- e1071::svm(class ~ x1 + x2, data = toy,
                       kernel = "linear", scale = FALSE)

# SVM decision boundary: w'x + b = 0  →  x2 = -(w1*x1 + b) / w2
w  <- t(mdl_lin$coefs) %*% mdl_lin$SV   # weight vector
b  <- -mdl_lin$rho                       # bias

slope     <-  -w[1] / w[2]
intercept <-  -b    / w[2]

# Margin lines: ±1/||w|| offset from decision boundary
margin_off <- 1 / w[2]

# Identify support vectors
sv_idx <- mdl_lin$index
sv_df  <- toy[sv_idx, ]

ggplot(toy, aes(x1, x2, colour = class, shape = class)) +
  geom_point(size = 3.5, alpha = 0.85) +
  # decision boundary
  geom_abline(slope = slope, intercept = intercept,
              colour = "black", linewidth = 1.1) +
  # margin boundaries
  geom_abline(slope = slope, intercept = intercept + margin_off,
              colour = "steelblue", linewidth = 0.8, linetype = "dashed") +
  geom_abline(slope = slope, intercept = intercept - margin_off,
              colour = "tomato",    linewidth = 0.8, linetype = "dashed") +
  # highlight support vectors
  geom_point(data = sv_df, aes(x1, x2),
             shape = 21, size = 6, stroke = 1.5,
             colour = "black", fill = NA) +
  annotate("text", x = -3.5, y =  2.5, label = "Margin boundary",
           colour = "steelblue", size = 3.5) +
  annotate("text", x =  0.5, y = -3,   label = "Margin boundary",
           colour = "tomato",    size = 3.5) +
  annotate("text", x = -0.8, y =  1.5, label = "Decision\nboundary",
           colour = "black",    size = 3.5) +
  annotate("text", x =  2.5, y = -2.5,
           label = "◯ = Support\n   Vectors",
           colour = "grey30", size = 3.2) +
  scale_colour_manual(values = c("Class A" = "steelblue",
                                 "Class B" = "tomato")) +
  scale_shape_manual(values  = c("Class A" = 16, "Class B" = 17)) +
  labs(title    = "SVM: Maximum-Margin Classifier",
       subtitle = "The dashed lines form the margin; circled points are support vectors",
       x = "Feature 1 (x₁)", y = "Feature 2 (x₂)",
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

---

# Core SVM Terminology

| Term | Plain-English Meaning |
|---|---|
| **Hyperplane** | The decision boundary (a line in 2-D, a plane in 3-D, …) |
| **Margin** | The gap between the two margin boundaries |
| **Support Vectors** | Data points closest to the boundary; they define the margin |
| **Hard Margin** | Allows zero misclassifications (only works if data is perfectly separable) |
| **Soft Margin (C)** | Allows some misclassifications; controlled by cost parameter **C** |
| **Kernel** | A mathematical trick that maps data into a higher-dimensional space |
| **γ (gamma)** | Controls how far the influence of a single training point reaches |

---

# Hard Margin vs. Soft Margin

Real data is rarely perfectly separable.  The **soft-margin SVM** introduces a
cost parameter **C**:

* **Large C** → penalise misclassifications heavily → narrow margin, low bias, high variance (overfitting risk)  
* **Small C** → tolerate some errors → wide margin, higher bias, lower variance (better generalisation)

```{r soft-margin, fig.cap="Fig 2 – Effect of the cost parameter C on the decision boundary."}
# Slightly overlapping data
set.seed(7)
noisy <- data.frame(
  x1    = c(rnorm(40, -1.5, 1.2), rnorm(40,  1.5, 1.2)),
  x2    = c(rnorm(40, -1.5, 1.2), rnorm(40,  1.5, 1.2)),
  class = factor(rep(c("A","B"), each = 40))
)

# Helper: fit SVM + return a tidy grid for plotting
plot_svm_boundary <- function(data, cost_val, title_str) {
  m <- e1071::svm(class ~ x1 + x2, data = data,
                  kernel = "linear", cost = cost_val, scale = FALSE)

  # prediction grid
  grid <- expand.grid(
    x1 = seq(min(data$x1)-1, max(data$x1)+1, length.out = 200),
    x2 = seq(min(data$x2)-1, max(data$x2)+1, length.out = 200)
  )
  grid$pred <- predict(m, grid)

  ggplot() +
    geom_tile(data = grid, aes(x1, x2, fill = pred), alpha = 0.25) +
    geom_point(data = data, aes(x1, x2, colour = class, shape = class),
               size = 2.5) +
    geom_point(data = data[m$index, ],
               aes(x1, x2), shape = 21, size = 5, stroke = 1.2,
               colour = "black", fill = NA) +
    scale_fill_manual(values   = c("A" = "steelblue", "B" = "tomato"),
                      guide = "none") +
    scale_colour_manual(values = c("A" = "steelblue", "B" = "tomato")) +
    scale_shape_manual(values  = c("A" = 16, "B" = 17)) +
    labs(title = title_str,
         x = "x₁", y = "x₂", colour = NULL, shape = NULL) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5))
}

p_small  <- plot_svm_boundary(noisy, 0.01, "Small C = 0.01\n(Wide margin, tolerates errors)")
p_medium <- plot_svm_boundary(noisy, 1,    "Medium C = 1\n(Balanced)")
p_large  <- plot_svm_boundary(noisy, 100,  "Large C = 100\n(Narrow margin, strict)")

gridExtra::grid.arrange(p_small, p_medium, p_large, ncol = 3)
```

---

# The Kernel Trick

When data is **not linearly separable** in its original space, a **kernel
function** implicitly maps it into a higher-dimensional space where a linear
hyperplane *can* separate the classes.

> You never actually compute the coordinates in the new space – the kernel
> evaluates dot products there cheaply.  This is the *kernel trick*.

## Common Kernels

| Kernel | Formula | Best for |
|---|---|---|
| **Linear** | $K(x,z)=x^Tz$ | Text, high-dimensional sparse data |
| **Polynomial** | $K(x,z)=(x^Tz+c)^d$ | Image classification |
| **RBF / Gaussian** | $K(x,z)=e^{-\gamma\|x-z\|^2}$ | General-purpose, non-linear problems |
| **Sigmoid** | $K(x,z)=\tanh(\alpha x^Tz+c)$ | Neural-network-like behaviour |

```{r xor-data}
# XOR-like data: cannot be separated by a straight line
set.seed(42)
n <- 60
xor_df <- data.frame(
  x1    = c(rnorm(n,  1.5, 0.5), rnorm(n, -1.5, 0.5),
             rnorm(n,  1.5, 0.5), rnorm(n, -1.5, 0.5)),
  x2    = c(rnorm(n,  1.5, 0.5), rnorm(n, -1.5, 0.5),
             rnorm(n, -1.5, 0.5), rnorm(n,  1.5, 0.5)),
  class = factor(rep(c("A","A","B","B"), each = n))
)
```

```{r kernels-comparison, fig.cap="Fig 3 – Linear vs. RBF kernel on non-linearly separable (XOR) data."}
plot_kernel <- function(data, kern, cost_val = 1, gamma_val = 1, title_str) {
  if (kern == "linear") {
    m <- e1071::svm(class ~ x1 + x2, data = data,
                    kernel = kern, cost = cost_val, scale = FALSE)
  } else {
    m <- e1071::svm(class ~ x1 + x2, data = data,
                    kernel = kern, cost = cost_val,
                    gamma = gamma_val, scale = FALSE)
  }
  grid <- expand.grid(
    x1 = seq(min(data$x1)-0.5, max(data$x1)+0.5, length.out = 300),
    x2 = seq(min(data$x2)-0.5, max(data$x2)+0.5, length.out = 300)
  )
  grid$pred <- predict(m, grid)

  ggplot() +
    geom_tile(data = grid, aes(x1, x2, fill = pred), alpha = 0.3) +
    geom_point(data = data, aes(x1, x2, colour = class, shape = class),
               size = 2, alpha = 0.9) +
    scale_fill_manual(values   = c("A" = "#4e9af1", "B" = "#f17a4e"),
                      guide = "none") +
    scale_colour_manual(values = c("A" = "#1a62c4", "B" = "#c43a1a")) +
    scale_shape_manual(values  = c("A" = 16, "B" = 17)) +
    labs(title = title_str, x = "x₁", y = "x₂",
         colour = NULL, shape = NULL) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold"))
}

pk_lin <- plot_kernel(xor_df, "linear", title_str = "Linear Kernel\n(Fails on XOR)")
pk_rbf <- plot_kernel(xor_df, "radial", gamma_val = 0.5,
                      title_str = "RBF Kernel\n(Handles non-linear boundary)")

gridExtra::grid.arrange(pk_lin, pk_rbf, ncol = 2)
```

---

# The Gamma (γ) Parameter

**γ** controls the *reach* of each training example in an RBF kernel:

* **High γ** → each point influences only its close neighbours → tight, complex boundary (overfitting risk)  
* **Low γ**  → each point influences a wide area → smooth, simpler boundary (underfitting risk)

```{r gamma-effect, fig.cap="Fig 4 – How gamma shapes the RBF decision boundary."}
make_gamma_plot <- function(gval, label) {
  m <- e1071::svm(class ~ x1 + x2, data = xor_df,
                  kernel = "radial", cost = 1,
                  gamma = gval, scale = FALSE)
  grid <- expand.grid(
    x1 = seq(-3.5, 3.5, length.out = 250),
    x2 = seq(-3.5, 3.5, length.out = 250)
  )
  grid$pred <- predict(m, grid)

  ggplot() +
    geom_tile(data = grid, aes(x1, x2, fill = pred), alpha = 0.3) +
    geom_point(data = xor_df, aes(x1, x2, colour = class, shape = class),
               size = 1.8, alpha = 0.85) +
    scale_fill_manual(values   = c("A" = "#4e9af1","B" = "#f17a4e"),
                      guide = "none") +
    scale_colour_manual(values = c("A" = "#1a62c4","B" = "#c43a1a"),
                        guide = "none") +
    scale_shape_manual(values  = c("A" = 16, "B" = 17), guide = "none") +
    labs(title = label, x = "x₁", y = "x₂") +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}

g1 <- make_gamma_plot(0.05, "γ = 0.05  (Smooth)")
g2 <- make_gamma_plot(0.5,  "γ = 0.5   (Balanced)")
g3 <- make_gamma_plot(5,    "γ = 5     (Complex)")
g4 <- make_gamma_plot(50,   "γ = 50    (Overfit)")

gridExtra::grid.arrange(g1, g2, g3, g4, ncol = 4)
```

---

# Fitting SVM with `caret`

The `caret` package provides a **unified interface** for hundreds of ML models.
We will use the classic **Iris dataset** (3 flower species, 4 features).

## Data Preparation

```{r data-prep}
data(iris)
str(iris)
summary(iris)
```

```{r data-split}
# 75/25 stratified split
train_idx <- createDataPartition(iris$Species, p = 0.75, list = FALSE)
train_data <- iris[ train_idx, ]
test_data  <- iris[-train_idx, ]

cat("Training rows:", nrow(train_data), "\n")
cat("Testing rows: ", nrow(test_data),  "\n")
```

## Define Cross-Validation

```{r cv-control}
# 10-fold cross-validation, repeated 3 times
ctrl <- trainControl(
  method          = "repeatedcv",
  number          = 10,     # 10 folds
  repeats         = 3,      # repeat 3 times
  classProbs      = TRUE,   # needed for ROC
  summaryFunction = multiClassSummary,
  savePredictions = "final"
)
```

## Train a Linear SVM

```{r linear-svm}
set.seed(42)
svm_linear <- train(
  Species ~ .,
  data      = train_data,
  method    = "svmLinear",     # linear kernel via kernlab
  trControl = ctrl,
  preProcess = c("center", "scale"),   # standardise features
  tuneGrid  = expand.grid(C = c(0.01, 0.1, 1, 10, 100))
)

svm_linear
plot(svm_linear, main = "Linear SVM – Accuracy vs. Cost (C)")
```

## Train an RBF (Radial) SVM

```{r rbf-svm}
set.seed(42)
svm_rbf <- train(
  Species ~ .,
  data      = train_data,
  method    = "svmRadial",     # RBF kernel via kernlab
  trControl = ctrl,
  preProcess = c("center", "scale"),
  tuneLength = 10              # auto-search C and sigma
)

svm_rbf
plot(svm_rbf, main = "RBF SVM – Accuracy vs. Cost (C)")
```

## Comparing the Two Models

```{r model-comparison, fig.cap="Fig 5 – Resampling accuracy distributions for Linear vs. RBF SVM."}
resamps <- resamples(list(Linear = svm_linear, RBF = svm_rbf))
summary(resamps)
bwplot(resamps, metric = "Accuracy",
       main = "Cross-Validation Accuracy: Linear vs. RBF SVM")
```

---

# Model Evaluation on the Test Set

```{r evaluation}
# Predictions
pred_linear <- predict(svm_linear, test_data)
pred_rbf    <- predict(svm_rbf,    test_data)

# Confusion matrices
cat("─── Linear SVM ───────────────────────────────\n")
cm_lin <- confusionMatrix(pred_linear, test_data$Species)
print(cm_lin)

cat("\n─── RBF SVM ──────────────────────────────────\n")
cm_rbf <- confusionMatrix(pred_rbf, test_data$Species)
print(cm_rbf)
```

```{r confusion-heatmap, fig.cap="Fig 6 – Confusion matrix heatmaps for both models."}
cm_to_df <- function(cm, model_name) {
  as.data.frame(cm$table) %>%
    mutate(Model = model_name)
}

cm_df <- bind_rows(
  cm_to_df(cm_lin, "Linear SVM"),
  cm_to_df(cm_rbf, "RBF SVM")
)

ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  facet_wrap(~Model) +
  labs(title = "Confusion Matrices – Test Set",
       x = "Actual Class", y = "Predicted Class", fill = "Count") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x  = element_text(angle = 30, hjust = 1),
        legend.position = "right")
```

---

# Decision Boundary Visualisation (2-D Slice)

Because we have 4 features, we visualise the boundary using only the two most
discriminating features: **Petal.Length** and **Petal.Width**.

```{r 2d-boundary, fig.cap="Fig 7 – SVM decision regions on Petal.Length vs. Petal.Width."}
# Re-fit on the 2-feature subset for visualisation
iris2 <- iris[, c("Petal.Length", "Petal.Width", "Species")]

svm2_lin <- e1071::svm(Species ~ ., data = iris2,
                        kernel = "linear", cost = 1, scale = TRUE)
svm2_rbf <- e1071::svm(Species ~ ., data = iris2,
                        kernel = "radial", cost = 1, gamma = 0.5, scale = TRUE)

# Prediction grid
x1_range <- seq(min(iris2$Petal.Length) - 0.3,
                max(iris2$Petal.Length) + 0.3, length.out = 300)
x2_range <- seq(min(iris2$Petal.Width)  - 0.3,
                max(iris2$Petal.Width)  + 0.3, length.out = 300)
grid2 <- expand.grid(Petal.Length = x1_range, Petal.Width = x2_range)

grid2$pred_lin <- predict(svm2_lin, grid2)
grid2$pred_rbf <- predict(svm2_rbf, grid2)

pal3 <- c("setosa"     = "#4e9af1",
           "versicolor" = "#f1c84e",
           "virginica"  = "#f17a4e")

plot_iris_boundary <- function(grid_col, title_str) {
  ggplot() +
    geom_tile(data = grid2,
              aes(x = Petal.Length, y = Petal.Width,
                  fill = .data[[grid_col]]),
              alpha = 0.35) +
    geom_point(data = iris2,
               aes(Petal.Length, Petal.Width,
                   colour = Species, shape = Species),
               size = 2.5, alpha = 0.9) +
    scale_fill_manual(values = pal3, guide = "none") +
    scale_colour_manual(values = pal3) +
    scale_shape_manual(values = c(16, 17, 15)) +
    labs(title = title_str,
         x = "Petal Length (cm)", y = "Petal Width (cm)",
         colour = "Species", shape = "Species") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5, face = "bold"))
}

p_lin2 <- plot_iris_boundary("pred_lin", "Linear SVM Decision Regions")
p_rbf2 <- plot_iris_boundary("pred_rbf", "RBF SVM Decision Regions")

gridExtra::grid.arrange(p_lin2, p_rbf2, ncol = 2)
```

---

# Feature Importance

SVM does not produce feature importance natively, but `caret` provides a
**variable importance** proxy based on the absolute values of the model weights
(linear kernel) or filter-based scores (non-linear).

```{r feature-importance, fig.cap="Fig 8 – Variable importance from the linear SVM."}
vi <- varImp(svm_linear, scale = TRUE)
plot(vi, main = "Variable Importance – Linear SVM (Iris)",
     col = "steelblue")
```

---

# Hyperparameter Tuning with a Custom Grid

```{r tuning-grid}
set.seed(42)
tune_grid <- expand.grid(
  sigma = c(0.01, 0.05, 0.1, 0.5),
  C     = c(0.1, 1, 10, 100)
)

svm_tuned <- train(
  Species ~ .,
  data       = train_data,
  method     = "svmRadial",
  trControl  = ctrl,
  preProcess = c("center", "scale"),
  tuneGrid   = tune_grid,
  metric     = "Accuracy"
)

svm_tuned$bestTune
```

```{r tuning-heatmap, fig.cap="Fig 9 – Accuracy heatmap across the C × σ grid."}
tune_res <- svm_tuned$results

ggplot(tune_res, aes(x = factor(C), y = factor(sigma), fill = Accuracy)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(Accuracy, 3)), size = 3.5, fontface = "bold") +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       limits = c(min(tune_res$Accuracy),
                                  max(tune_res$Accuracy))) +
  labs(title = "RBF SVM Tuning – Accuracy Grid",
       x = "Cost (C)", y = "Sigma (σ)", fill = "CV Accuracy") +
  theme_minimal(base_size = 13)
```

---

# Learning Curves

Learning curves help diagnose whether a model suffers from **high bias**
(underfitting) or **high variance** (overfitting).

```{r learning-curve, fig.cap="Fig 10 – Learning curves for the RBF SVM."}
# Compute accuracy at different training-set sizes
sizes    <- seq(0.10, 0.90, by = 0.10)
lc_ctrl  <- trainControl(method = "cv", number = 5)

lc_results <- lapply(sizes, function(p) {
  idx    <- createDataPartition(train_data$Species, p = p, list = FALSE)
  subset <- train_data[idx, ]
  m <- train(Species ~ ., data = subset, method = "svmRadial",
             trControl = lc_ctrl, preProcess = c("center","scale"),
             tuneGrid  = data.frame(sigma = svm_tuned$bestTune$sigma,
                                    C     = svm_tuned$bestTune$C))
  train_acc <- max(m$results$Accuracy)
  test_acc  <- mean(predict(m, test_data) == test_data$Species)
  data.frame(size = nrow(subset), train_acc, test_acc)
})

lc_df <- bind_rows(lc_results) %>%
  pivot_longer(c(train_acc, test_acc),
               names_to = "set", values_to = "accuracy") %>%
  mutate(set = recode(set,
                      train_acc = "Training",
                      test_acc  = "Validation"))

ggplot(lc_df, aes(size, accuracy, colour = set, group = set)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("Training" = "steelblue",
                                 "Validation" = "tomato")) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0.7, 1.02)) +
  labs(title    = "Learning Curves – RBF SVM (Iris)",
       subtitle = "Converging curves indicate a well-fitted model",
       x = "Training Set Size", y = "Accuracy",
       colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

---

# SVM for Regression (SVR)

SVM can also solve **regression** problems.  Instead of a margin that separates
classes, it tries to fit as many points as possible inside an **ε-tube** around
the predicted function.

```{r svr, fig.cap="Fig 11 – Support Vector Regression on simulated non-linear data."}
set.seed(1)
n_reg   <- 100
x_reg   <- seq(0, 4*pi, length.out = n_reg)
y_reg   <- sin(x_reg) + rnorm(n_reg, 0, 0.25)
reg_df  <- data.frame(x = x_reg, y = y_reg)

svr_mdl <- e1071::svm(y ~ x, data = reg_df,
                       kernel = "radial", epsilon = 0.1,
                       cost = 10, gamma = 0.3)

pred_svr  <- predict(svr_mdl, reg_df)
sv_reg_df <- reg_df[svr_mdl$index, ]

ggplot(reg_df, aes(x, y)) +
  geom_point(alpha = 0.5, colour = "grey50", size = 2) +
  geom_point(data = sv_reg_df, aes(x, y),
             shape = 21, size = 4, stroke = 1.2,
             colour = "tomato", fill = NA) +
  geom_line(aes(y = pred_svr), colour = "steelblue",
            linewidth = 1.2) +
  geom_ribbon(aes(ymin = pred_svr - 0.1,
                  ymax = pred_svr + 0.1),
              alpha = 0.15, fill = "steelblue") +
  labs(title    = "Support Vector Regression (SVR)",
       subtitle = "Blue line = prediction; shaded band = ε-tube; red circles = support vectors",
       x = "x", y = "y") +
  theme_minimal(base_size = 13)
```

---

# Summary Cheat-Sheet

```{r summary-table, echo=FALSE}
summary_df <- data.frame(
  Concept           = c("Hyperplane","Margin","Support Vectors",
                        "Cost C","Kernel","Gamma γ",
                        "Hard Margin","Soft Margin","SVR"),
  `One-liner`       = c(
    "Decision boundary separating classes",
    "Distance between boundary and nearest points",
    "Points on the margin edges that define the boundary",
    "Penalty for misclassifications; higher C = stricter",
    "Maps data to higher dimensions for non-linear separation",
    "Controls reach of each training point in RBF kernel",
    "No errors allowed; requires perfectly separable data",
    "Allows some errors; controlled by C",
    "SVM adapted for predicting continuous values"
  ),
  `Rule of Thumb` = c(
    "Choose kernel based on data geometry",
    "Wider = better generalisation",
    "Fewer SVs = simpler model",
    "Tune via cross-validation",
    "RBF is a safe default; linear for high-dim text",
    "Start with γ = 1/n_features",
    "Rarely used in practice",
    "Default choice for real data",
    "Use ε-SVR; tune ε and C"
  )
)
knitr::kable(summary_df, align = "lll",
             caption = "SVM Concept Cheat-Sheet")
```

---

# Session Info

```{r session}
sessionInfo()
```
